Structuring Theories with Implicit Morphisms

### 2018-10-14 appl.tex
some clean up

\begin{example}
	Consider a variation of the theory $\cn{Monoid}$ from \autoref{syn:incl} in a different library:
	\begin{mmtcode}
Monoid2 =
  M: type
  connective : M ⟶ M ⟶ M	 # 1 ∘ 2 
  neutral	: M
\end{mmtcode}
This theory is isomorphic to the previously introduced theory $\cn{Monoid}$ under the trivial renaming
	\begin{mmtcode}
Monoids : Monoid2 -> Monoid =
  M 		:= U
  connective 	:= op 
  neutral	:= unit
\end{mmtcode}
\end{example}

\paragraph{Definitional Extensions}
We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
\begin{compactitem}
 \item only constant declarations with definiens,
 \item only include declarations of theories that are definitional extensions of $S$.
\end{compactitem}
\begin{example}
	Imagine extending the theory $\cn{Group}$ from \autoref{syn:incl} by provable theorems:
		\begin{mmtcode}
Group_Theorems =
  include Group
  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)
\end{mmtcode}
\end{example}

@@

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
\end{compactitem}
\begin{example}
	We adapt our theory $\cn{Group\_Theorems}$ with the newly introduced keyword, which allows us to extend $\cn{Group}$ directly while having the new theorem $\cn{inverse\_idem}$ available:
		\begin{mmtcode}
Group_Theorems =
  definitional include Group
  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)

More_Theorems =
  include ?Group
  some_other_theorem : ⊦ ??	= (proof using inverse_idem)
\end{mmtcode}
\end{example}

@@

\begin{example}\label{group:iso}
Having declared the morphism $\cn{DG2G}$ (as in \autoref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
\begin{mmtcode}
implicit G2DG : Group -> DivisionGroup =
  U := U
  op := [a,b] a / ((unit/unit) / b) 
  unit := unit/unit
  inverse := [a] (unit/unit) / a
\end{mmtcode}
\end{example}

@@

With implicit morphisms, we can solve this problem by making only the following local changes:
\begin{compactenum}
  \item We rename $s$ to $s'$.
  \item We delete the declaration $n:E$ from $s'$.
  \item We create new theories $r'=\{\icl{r},\,D\}$ and $s=\{\icl{s'},D\}$.
  \item We change $t$ to $t=\{\icl{r'},\icl{s'}\}$.
  \item We add an implicit morphism $s\to t$ that maps $s?n$ to $r'?n$.
\end{compactenum}
Now $t$ has the desired new structure.
But, all old references to $s?n$ stay well-formed so that no global changes are needed, as in \autoref{fig:refac}.

\begin{figure}\begin{center}

\begin{minipage}[t]{0.4\textwidth}
\begin{center}
Before:

\begin{tikzpicture}[scale=1.5]
  \node (t) at (0,0) {$t$};
  \node (r) at (-1,-2) {$r$};
  \node (s) at (1,-1) {$s$};
  \draw[right hook-latex] (r) to (t) {};
  \draw[right hook-latex] (s) to (t) {};

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}
\end{center}
\end{minipage}
\begin{minipage}[t]{0.4\textwidth}
\begin{center}
After:

\begin{tikzpicture}[scale=1.5]
  \node (t) at (0,0) {$t$};
  \node (rp) at (-1,-1) {$r'$};
  \node (s) at (1,-1) {$s$};

@@

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}
\end{center}
\end{minipage}

\end{center}
\caption{An Example for Transparent Refactoring}\label{fig:refac}
\end{figure}

>>>>

\begin{example}[Renaming]
Consider a variation of the theory $\cn{Monoid}$ from Ex.~\ref{syn:incl} in a different library:
\begin{mmtcode}
Monoid2 =
  M: type
  connective : M ⟶ M ⟶ M	 # 1 ∘ 2 
  neutral	: M
\end{mmtcode}
This theory is isomorphic to the previously introduced theory \cn{Monoid} under the trivial renaming
	\begin{mmtcode}
renaming MonoidRen : Monoid2 -> Monoid =
  M 		:= U
  connective 	:= op 
  neutral	:= unit
\end{mmtcode}
\end{example}

\paragraph{Definitional Extensions}
We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
\begin{compactitem}
 \item only constant declarations with definiens,
 \item only include declarations of theories that are definitional extensions of $S$.
\end{compactitem}

@@

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
\end{compactitem}

\begin{example}[Extension with a Theorem]
We extend the theory $\cn{Group}$ from Ex.~\ref{syn:incl} with a theorem
\begin{mmtcode}
definitional InverseInvolution =
  include Group
  inverse_invol : ∀[x] (x⁻¹)⁻¹≐x 	= (proof omitted)
\end{mmtcode}
\end{example}

@@

\begin{example}\label{group:iso}
Having declared the morphism $\cn{DG2G}$ (as in Ex.~\ref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
\begin{mmtcode}
implicit G2DG : Group -> DivisionGroup =
  U := U
  op := [a,b] a/(unit/b) 
  unit := unit
  inverse := [a] unit/a
\end{mmtcode}
\end{example}

@@

With implicit morphisms, we can solve this problem by making only the following local changes:
\begin{compactenum}
  \item We rename $s$ to $s'$.
  \item We delete the declaration $n:E$ from $s'$.
  \item We create new theories $r'=\{\icl{r},\,D\}$ and $s=\{\icl{s'},D\}$.
  \item We change $t$ to $t=\{\icl{r'},\icl{s'}\}$.
  \item We add an implicit morphism $s\to t$ that maps $s?n$ to $r'?n$.
\end{compactenum}
Now $t$ has the desired new structure.
But, all old references to $s?n$ stay well-formed so that no global changes are needed, as in Fig.~\ref{fig:refac}.

The situation before (left) and after (right) refactoring is given below:
\begin{center}
\begin{tikzpicture}[scale=1]
  \node (t) at (0,0) {$t$};
  \node (r) at (-1,-2) {$r$};
  \node (s) at (1,-1) {$s$};
  \draw[right hook-latex] (r) to (t) {};
  \draw[right hook-latex] (s) to (t) {};

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}
\tb
\begin{tikzpicture}[scale=1]
  \node (t) at (0,0) {$t$};
  \node (rp) at (-1,-1) {$r'$};
  \node (s) at (1,-1) {$s$};

@@

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}
\end{center}

### 2018-10-14 example.tex
some clean up

Implicit morphisms provide a simple solution to this problem because they behave effectively like inclusions but can be added later on.
In the above example, we would
\begin{compactitem}
 \item define \cn{CancellationMonoid} with an include from \cn{Monoid},
 \item keep \cn{Group} as it is, i.e., also with an include \cn{Monoid},
 \item add an implicit morphism $\cn{CancellationMonoid}\to\cn{Group}$.
\end{compactitem}

In the sequel, we develop a similar, more complex example: we build a hierarchy of theories between \cn{Band} and \cn{Semilattice}.
\ednote{@DM: this is not quite the lattice from the Wikipedia page: it consists of the 3x3 square plus Bands at the top}
All of these theories are of the form $t=\{\icl{\cn{Band}},\,a:\vdash F\}$, where $F$ is a universally quantified equational axiom.\footnote{All varieties of bands can be axiomatized in this way.}
In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a:\vdash \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.
Fig.~\ref{fig:bands} shows these theories.

There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.\ednote{@DM: no need to give the morphisms, but link to the online resource if you have time to work them all out}
We make all of these morphisms implicit.
It is straightforward to prove that the diagram commutes: any two morphisms are identical except for assignment for the axiom.
Here, \mmt can discharge the proof obligation easily using proof irrelevance.

>>>>

Implicit morphisms provide a simple solution to this problem because they behave effectively like inclusions but can be added later on.
In the above example, we would
\begin{compactitem}
 \item define \cn{CancellationMonoid} with an include from \cn{Monoid},
 \item keep \cn{Group} as it is, i.e., also with an include \cn{Monoid},
 \item add an implicit morphism $\cn{CancellationMonoid}\to\cn{Group}$.
\end{compactitem}
\medskip

As a more complex example, we have built a hierarchy of theories between \cn{Band} and \cn{Semilattice}.
\ednote{@DM: this is not quite the lattice from the Wikipedia page: it consists of the 3x3 square plus Bands at the top}
All of these theories are of the form $t=\{\icl{\cn{Band}},\,a:\vdash F\}$, where $F$ is a universally quantified equational axiom.\footnote{All varieties of bands can be axiomatized in this way.}
In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a:\vdash \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.
Fig.~\ref{fig:bands} shows these theories.

There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.%
\footnote{Our formalization of the lattice of bands using implicit morphisms can be found at \url{https://gl.mathhub.info/MitM/smglom/blob/devel/source/algebra/bands.mmt}.}
We make all of these morphisms implicit.
It is straightforward to prove that the diagram commutes: any two morphisms are identical except for assignment for the axiom.
Here, \mmt can discharge the proof obligation easily using proof irrelevance.

### 2018-10-14 mmt.tex
some clean up

The declarations in theories and morphisms are subject to typing conditions that are not essential for our purposes here.
We only mention the main theorem that guarantees that well-typed morphisms preserve judgments, i.e., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
This includes the preservation of truth via the propositions-as-types principle where $E$ is a proposition and $e$ its proof.

@@

\paragraph{Well-Formed Expressions}
The inference system to define well-formed expressions is not a primary interest of this paper.
Intuitively, the set $\Exp{T}$ of expressions over $T$ includes all syntax trees that can be formed using the constants from $\flt{T}$.
Moreover, well-formed modules may only use expressions that satisfy certain typing and equality judgments.
We refer to \cite{rabe:howto:14} for details.

The only aspect of the inference system for these judgments that is relevant to our purposes here is the rules for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
\footnote{The \mmt tool accepts $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition. In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$. This is important for efficiency but not essential for our purposes here.}

@@

\paragraph{Logics in \mmt}
\ednote{explain that we use a fixed instance of MMT which adds LF and FOL; only relevant for examples}

@@

\begin{example}\label{ex:thgroup}
The (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
\begin{mmtcode}
Group =
  U	: type
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  inverse : U ⟶ U   # 1 ⁻¹ 
  // axioms omitted
\end{mmtcode}
\begin{mmtcode}
DivisionGroup =
  U : type
  div : U ⟶ U ⟶ U  # 1 / 2 
  unit : U
  // axioms omitted
\end{mmtcode}
Here we use \lstinline|# <not>| to indicate notations; these are part of the language supported by \mmt but omitted from the formal grammar above because we only need them here to present legible examples.

Moreover, we use a few constants from a type systems\ednote{finish}$\forall,\doteq$ etc.
\end{example}

>>>>

@@

\paragraph{Logics and Well-Formed Expressions}
The logic and the definition of well-formed expressions are not a primary interest of this paper, and we only recap the essential structure needed in the sequel.
We refer to \cite{rabe:howto:14} for details.

\mmt itself is independent of the base logic and provides a theory and theory morphism layer on top of an arbitrary declarative language.
Our results in this paper are similarly logic-independent.
The syntax and semantics of that base logic is provided by a set of rules for the judgments $\vdash_T e:E$ and $\vdash_T e=e'$ for $e,e',E\in\Exp{T}$.
The declarations in theories and morphisms are subject to typing conditions using these judgments.
The main theorem about \mmt is that well-typed morphisms $M:S\to T$ preserve all judgments, e.g,., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
This includes the preservation of truth via the propositions-as-types principle if $E$ is a proposition and $e$ its proof.

The only aspect of the inference system for these judgments that is relevant to our purposes here is the typing rule for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
\footnote{The \mmt tool accepts $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition. In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$. This is important for efficiency but not essential for our purposes here.}

@@

@@

\begin{example}\label{ex:thgroup}
For the purposes of our running examples, we assume a fixed base logic that provides a first-order logic a simple type system.
\lstinline|type| is the universe of types, \lstinline|A ⟶ B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
We also use \mmt notations, which are attached to constant declarations as \lstinline|# <not>|; these are omitted from the formal grammar above because we only use them in the examples.

Then the (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
\begin{mmtcode}
Group =
  U		: type
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  inverse 	: U ⟶ U   # 1 ⁻¹ 
  // axioms omitted
\end{mmtcode}
\begin{mmtcode}
DivisionGroup =
  U 	: type
  div 	: U ⟶ U ⟶ U  # 1 / 2 
  unit 	: U
  // axioms omitted
\end{mmtcode}
\end{example}
